Formulating the posterior for \tardis

### 2015-09-28 posterior/model.tex
[posterior] Rewrite `incorporating the observations`
Math didn't really change but I added a lot explanation due to comments
from Hans Eggers

Let \Lumi{} denote the predicted or true luminosity in bin $i$ and
$\vecL$ the ordered collection of luminosities. I assume from a
telescope, we get the probability of the data $\mathcal{D}$ given
$\vecth$ only as a function of $\vecL$ where the data are
implicit---we don't have them---and no correlation between bins is
given or assumed. The maximum-entropy distribution given only the
means and standard deviations $\expin, i=1\dots n_b$ is the normal or
Gaussian distribution, so we set
\begin{equation}
  \label{eq:like-L}
  P(\mathcal{D} | \vecL) = \prod_{i=1}^{n_b} \gaussian\given{\Lumi}{\mu_i, \sigma_i^2}.
\end{equation}

In principle, $\vecL(\vecth)$ is a deterministic function of
$\vecth$. But we are unable to calculate it as is; that's why we have
to estimate it from the \tardis{} simulation resulting in a
distribution $P\given{\vecL}{\vecth}$. To use Bayes'
theorem~\refeq{Bayes-full}, we need $P\given{\data}{\vecth}$ but we
only have $P\given{\data}{\vecL}$. The link between these three
densities is the law of total probability:
\begin{align}
  \label{eq:like-expanded}
  P\given{\data}{\vecth} &= \int \rmdx{\vecL} P\given{\vecL}{\vecth} P\given{\data}{\vecL}.
\end{align}
Note that in principle it should say $P\given{\data}{\vecL, \vecth}$
in \refeq{like-expanded} but in our case $\vecth$ doesn't appear
explicitly, so we can just plug in
$P\given{\data}{\vecL}$ from~\refeq{like-L}.

To complete the specification of the likelihood, we need to formulate
$P\given{\vecL}{\vecth}$; i.e., what we learn about the luminosities
in each bin from \tardis. A simplified account of how \tardis{} is run
follows, then I will show how to phrase this in statistical terms.

>>>>

Let \Lumi{} denote the true (and unknown) luminosity in bin $i$ and
and let $\vecL$ represent the ordered collection of luminosities. At a
telescope, a possible realization of what they can observe (charges in
CMOS sensor?) is $X$. The physicists operating the telescope have a
model how the luminosities $\vecL$ and other parameters of the
telescop, call them $\vecnu$, affect the outcome $X$. Quantitatively,
in their model they write down $P \given{X}{\vecL, \vecnu}$.  We
further assume that the telescope people average out the effect of
their nuisance parameters to report to us
\begin{align}
  \label{eq:aver-like}
  P \given{X}{\vecL} = \int \rmdx{\vecnu} P \given{X}{\vecL, \vecnu} P(\vecnu) .
\end{align}
Once the telescope performs the measurement, they actually observe
$X=\data$ and they tell us that $P \given{X=\data}{\vecL}$---viewed as
a function of $\vecL$---can be approximated by a product of Gaussians
characterized by the numbers $\expin, i=1 \dots n_b$:
\begin{equation}
  \label{eq:like-L}
  P(X=\mathcal{D} | \vecL) \propto \prod_{i=1}^{n_b} \gaussian\given{\Lumi}{\mu_i, \sigma_i^2}.
\end{equation}
Note that for a different realization of $X$, $ P \given{X}{\vecL}$
might need to be approximated by a functional form different from a
product Gaussian. But this is of no concern to us because we only care
about what was actually observed; i.e., $X=\data$. The numbers
$\expin$ of course depend on $X$ (and the effect of $\vecnu$!). There
is a $\propto$ sign in \refeq{like-L} because the lhs sign is a
probability density for $X$ but not for $\vecL$. We don't know what
$\data$ actually is: the telescope people don't tell us and we don't
need to know as long as we only care about the dependence on
$\vecL$. For our purposes, all we need are the values $\expin$ to plug
into \refeq{like-L} under the given assumptions. We can ignore the
normalization unless we compare different models in \tardis{}
that give rise to different normalization constants.

To use Bayes' theorem~\refeq{Bayes-full}, we need
$P\given{\data}{\vecth}$ but we only have $P\given{\data}{\vecL}$. The
link is the law of total probability:
\begin{align}
  \label{eq:like-expanded}
  P\given{X}{\vecth} &= \int \rmdx{\vecL} P\given{X}{\vecL, \vecth} P\given{\vecL}{\vecth} .
\end{align}
We \emph{assume} that our parameters of interest, $\vecth$, do \emph{not}
affect $X$ in any other way than through $\vecL$; i.e., $\vecth$
doesn't appear explicitly
\begin{align}
  \label{eq:no-expl-L}
  P\given{X}{\vecL, \vecth} = P\given{X}{\vecL}.
\end{align}
and we can reuse \refeq{aver-like} on the rhs.  In principle, $\vecL$
is a deterministic function of $\vecth$, $\vecL = f(\vecth)$. But we
are unable to simply evaluate this function; that's why we have to
estimate it from the \tardis{} simulation resulting in a distribution
$P\given{\vecL}{\vecth} \ne \delta(\vecL - f(\vecth))$. Specifically,
the outcome of \tardis{} depends on the random number seed and the
fact that only a finite number of (effective) photons are
simulated. In the limit of infinite number of photons,
$P\given{\vecL}{\vecth} \to \delta(\vecL - f(\vecth))$. Note that this
does not mean that we know the true value of the luminosity in bin $i$
for a given supernova. Instead it means that there is no uncertainty
due to the finite simulation size. Whatever \tardis{} spits out
depends on the physics assumptions coded up in \tardis{} and there can
still be systematic shifts.

Putting together equations~\refeq{like-L}--\refeq{no-expl-L}, the
likelihood is
\begin{align}
  \label{eq:like-summary}
  \likelihood(\vecth) \propto \int \rmdx{\vecL} \prod_{i=1}^{n_b}
  \gaussian\given{\Lumi}{\mu_i, \sigma_i^2} P\given{\vecL}{\vecth}
\end{align}
To complete the specification of the likelihood, we need to formulate
$P\given{\vecL}{\vecth}$; i.e., what we learn about the luminosities
in each bin from \tardis. A simplified account of how \tardis{} is run
follows, then I will show how to phrase this in statistical terms.
